The Micro Booster Neutrino Experiment (MicroBooNE) is a Liquid Argon Time Projection Chamber (LArTPC) designed for short-baseline neutrino physics at the Fermi National Accelerator Laboratory. The main physics goal of MicroBooNE is to address the low-energy excess of electron-like events observed by the MiniBooNE experiment and, if confirmed, clarify its nature.
The MiniBooNE experiment is a Cherenkov detector and this technology does not allow to distinguish between electrons and single photons in the final state. 
LArTPC detectors, instead, offer excellent granularity and powerful separation between electrons and photons. For this reason, they represent an ideal technology for the detection of electron neutrino interactions.
This thesis presents the first fully-automated electron neutrino selection in MicroBooNE. The selection looks for charged-current electron neutrino interactions with no pions and at least one proton in the final state. It is applied on a sub-sample of the data acquired by the detector corresponding to $4.34\times10^{19}$~protons-on-target in the Booster Neutrino Beam. A validation of the analysis is performed on two orthogonal side-bands, enriched with neutral-current and charged-current muon neutrinos interactions, respectively. The uncertainties of the neutrino cross-section, flux, and detector simulation are evaluated.
The MicroBooNE detector is placed off-axis with the Neutrinos at the Main Injector (NuMI) beam. An independent dataset of events acquired by triggering on the NuMI beam is used to measure the significance of the detection of electron neutrinos in the beam using the selection presented here. 
The sensitivity of the MicroBooNE experiment to the MiniBooNE low-energy excess of electron-like events is evaluated. The efficiency and background-rejection power necessary to achieve $5\sigma$ sensitivity are also quantified.